---
title: "How to generate a binned line plot from flow cytometry data using R"
output: 
  html_document: 
    highlight: tango
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(dplyr)
library(gridExtra)
library(grid)
library(lattice)
```

##The problem with the dot plot

Data produced from a dual-color flow cytometry experiment typically look like the scatter plots below, where one dot represents one cell, and x- and y-axes represent features of interest within each cell. 

```{r generate scatter plot, echo=FALSE, message=FALSE, warning=FALSE}
dat1 = read.csv('control.csv')
ggplot(dat1, aes(x = log10(dat1$FITC.A), y = log10(dat1$APC.A))) + 
  geom_bin2d(bins = 150) +
  ylim(0, 5) +
  xlim(0, 5) +
  xlab('feature 1') +
  ylab('feature 2') -> p1

dat2 = read.csv('experimentalSample1.csv')
ggplot(dat2, aes(x = log10(dat2$FITC.A), y = log10(dat2$APC.A))) +
  geom_bin2d(bins = 150) +
  ylim(0, 5) +
  xlim(0, 5) +
  xlab('feature 1') +
  ylab('feature 2') -> p2

dat3 = read.csv('experimentalSample2.csv')
ggplot(dat3, aes(x = log10(dat3$FITC.A), y = log10(dat3$APC.A))) +
  geom_bin2d(bins = 150) +
  ylim(0, 5) +
  xlim(0, 5) +
  xlab('feature 1') +
  ylab('feature 2') -> p3

grid.arrange(p1, p2, p3, ncol = 2)
```

While this can be a helpful way to understand the relationship between the two features within a cell, it's difficult to interpret small changes between samples with side-by-side scatter plots. 

To facilitate comparison of multiple samples by plotting them on a single graph, I composed the code below to import, clean, and plot data accordingly.

## But first, libraries

ggplot2 and dplyr libraries are from the 'Tidyverse' package in R and will be necessary later when we plot the data.

```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2) # generates different types of plots
library(dplyr) # facilitates manipulation of dataframes
```

## Importing the data

Here, I assume that the data are already in CSV format. (If not, you might have to open your FCS file in FlowJo or FACSDiva software and export it to a CSV for this step!)

```{r import}
# assign the data within your csv to an intuitive variable name
control = read.csv('control.csv') # negative control data that should not show signal in the y axis
protein1 = read.csv('experimentalSample1.csv') # experimental sample 1
protein2 = read.csv('experimentalSample2.csv') # experimental sample 2
```

## Parameters typically recorded in each flow experiment 

Those familiar with flow cytometry will recognize the column headings below. 

```{r view head of protein2 dataset}
# view a sample dataset
head(protein2)
```

For the uninitiated, the labels translate to the following cellular features:

* __FSC__ = forward scatter, which tells us about each cell's size
* __SSC__ = side scatter, which tells us about each cell's granularity
* __APC__ = some feature of interest that the experimenter has chosen to make detectable using a fluorescent marker, which emits signal at ~660 nm
* __FITC__ = same as APC, except its emission wavelength is around ~520 nm 
* __Time__ = The time (in seconds) each cell passed through the laser, generating the data above

* __.A__ refers to the area under the fluorescent curve generated from each event (i.e. each cell) and tells us how intense the signal for each parameters was 
* __.W__ refers to the width of the fluorescent curve generate from each event and plays a role in what's called a "gating" process, which isn't covered here 

##Subsetting the data of interest

My experimental question was: Relative to the control protein ("control"), does protein 1 ("protein1") or protein 2 ("protein2") induce more proliferative signal within each cell?

As alluded to above, I'm particularly interested in the APC and FITC columns. In my case, FITC measurements tell me how abundant a protein of interest is in each cell. A more intense FITC reading therefore means that cell contains a higher number of that specific protein. 

The APC-labelled feature provides information about each cell's intracellular signaling activity. An intense reading here means that the cell is receiving a lot of signal to divide and proliferate (think cancer).

Now that the data are imported, I want to extract the columns containing experimental information: 

```{r generate dataframe containing only data of interest: FITC.A, APC.A}
# generates a new variable containing a dataframe of only FITC and APC columns
controlDat = subset(control, select = c('FITC.A', 'APC.A'))
protein1Dat = subset(protein1, select = c('FITC.A', 'APC.A'))
protein2Dat = subset(protein2, select = c('FITC.A', 'APC.A'))
```

##Clean it up

After subsetting the features of interest into a new dataframe, I want to view summary statistics of each parameter in my experiment to get an idea of the quality of the data.

```{r view summary statistics of protein2 dataset}
# compute summary statistics of each column within the specified sample dataframe
summary(protein2Dat) 
```

In evaluating these statistics, I can see that both FITC and APC data 'min' values are negative, implying the datasets contain negative, and possibly 0 values. Because flow data are generally reported on a log scale, I want to clean the data of these values to prevent generating an error when applying a log transform: 

_Note: there are a number of ways to handle NA values, including imputation. However, removal is sufficient for my purposes._ 

```{r replace neg and 0 values with NA, remove rows containing NA}
# combine dataframes of the three samples into a list
datasets <- list(control = controlDat, protein1 = protein1Dat, protein2 = protein2Dat)

# generate a new variable where negative and 0 values within each dataframe in datasets have been removed
datasets.clean <- lapply(datasets, function(dataset) { 
  dataset[dataset <= 0] <- NA 
  dataset <- na.omit(dataset) 
})
```

###Double check

I confirm that only non-negative data remain by again viewing summary statistics of the three samples:

```{r}
# compute example of summary statistics to confirm removal of negative/0 data
summary(datasets.clean[["protein2"]]) 
```

##Transform the data

Minimum values of both FITC and APC data are now positive numbers!

In the next step, I will log transform each value, as that is the scale I want to use when visualizing the data; this is standard for flow cytometry data:

```{r log transform}
# calculates log10 of each value in the current dataset
datasets.clean.log <- lapply(datasets.clean, function(dataset) { 
  log10(dataset) 
})
```

##A function to bin values along the x axis and calculate the mean of corresponding y axis values

Now that the data have been transformed, it's time to bin the values and calculate the means of each bin. 

Because I want to know how protein abundance (FITC data, x-axis) relate to intracellular activity (APC data, y-axis), I will bin the values according to FITC intensity and calculate the mean of the corresponding APC values using a custom function:

_Note: the current version of this function does not include error handling! To successfully implement the code below on your data, ensure the 'dataset' passed into the function is a dataframe where desired x-axis data are in column 1 and desired y-axis data are in column 2._
```{r bin data according to x intensity, compute mean of corresponding y values}
  binnedMeans <- function(dataset, binwidth) {
    
    # assign column 1 data to x; column 2 data to y
    x <- dataset %>% select(1) 
    y <- dataset %>% select(2) 
    
    # generate vector of values from 0 to maximum x  
    bins <- seq(from = 0, to = max(x), by = binwidth); 
    
    # initiate empty vectors into which values calculated within for loop below will be passed
    x_bin <- vector("numeric", 100);   
    x_n <- vector("numeric", 100); 
    y_means <- vector("numeric", 100);  
    y_SEMs <- vector("numeric", 100); 
    
    count <- 1

    for(bin in bins) { 
  
      # locate and count x values within specified bin
      x_values <- x > bin & x <= (bin + binwidth); 
      x_n[count] <- sum(x_values); 
  
      # calculate and store mean of y values corresponding to 'x_values'
      y_mean <- mean(y[x_values]);
      y_means[count] <- y_mean; 
      
      # calculate and store SEM of y values corresponding to 'x_values'
      y_SEM <- sd(y[x_values]) / sqrt(sum(x_values)); 
      y_SEMs[count] <- y_SEM;
      
      # record bin + binwidth for use in plotting
      x_bin[count] <- bin + binwidth; 
      
      bin <- bin + binwidth; 
      count <- count + 1; 
    }
    # generate dataframe with values calculated above, remove any rows containing '0'
    output <- data.frame(x_bin, x_n, y_means, y_SEMs) 
    output$y_means[output$y_means == 0] <- NA; 
    final.output <- output[!is.na(output$y_means), ]
    return(final.output)
}
```

###Execute the new function on three dataframes

Having prepared the function above, I will apply it to each of the datasets of interest using lapply: 

```{r apply function}
# apply function written above to each sample dataframe
binned.data <- lapply(datasets.clean.log, function(dataset) {
    binnedMeans(dataset, 0.5)
})
```

##Convert the data to long format for dplyr and ggplot manipulation

Now that we have the infomation we want to plot, time to get it easily plottable (pretty sure that's a word) format.

```{r mutate Sample}
# add column to each dataset specifying sample
controlDatFinal <- mutate(binned.data[["control"]], Sample = "Control")
protein1DatFinal <- mutate(binned.data[["protein1"]], Sample = "Protein 1")
protein2DatFinal <- mutate(binned.data[["protein2"]], Sample = "Protein 2")

# concatenate data into single dataframe
longDat <- rbind(controlDatFinal, protein1DatFinal, protein2DatFinal)
```

##Visualize 

Time to plot our results!:

```{r plot}
# order samples as desired to appear in legend
longDat$Sample <- factor(longDat$Sample, levels = c("Protein 2", "Protein 1", "Control"))

# specify colors to be assigned to samples, ordered as specified above
sampleColors = c("steelblue4", "steelblue3", "steelblue1")

# plot the data
 ggplot(data = longDat, aes(x = x_bin, y = y_means, color = Sample)) +  
   scale_color_manual(values = sampleColors) +
   geom_errorbar(aes(ymin = y_means - y_SEMs, ymax = y_means + y_SEMs), width = .1) +  
   geom_line(aes(group = Sample)) + geom_point(size = 0.5) +
   ggtitle("The effect of protein abundance on intracellular signaling activity") +
   xlim(0, 5.5) +
   ylim(1.5, 3) + 
   xlab("Protein Abundance") +
   ylab("Intracellular Signaling Activity") 
```

Voilà! Now you can easily see the difference in these three samples using a single graph! 

The control looks like it doesn't "activate" intracellular signaling, whereas the experimental proteins are able to induce signaling within a certain range. 